---
title: "An introduction to spatial deformations using B-splines"
author: "Guilherme Ludwig, Ronaldo Dias, Paul Sampson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Introduction to spatial deformations using B-splines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette describes how to install and use the `bsplinedef` package, a package for finding spatial deformations using the tensor product of B-splines. This is a work in progress and subject to change. For now, the package fits the semi-parametric covariance model \[\Gamma(x_1, x_2) =  \sigma^2(\|f_1(x_1,x_2) - f_2(x_1,x_2)\|; \boldsymbol\Delta)\] where $x_1, x_2 \in G \subset \mathbb{R}^2$ are sampled spatial locations in a geostatistical domain $G;$ $f_1,$ $f_2$ are the tensor product of B-spline functions, each with $K_1 \times K_2$ basis functions, respectively; and $\sigma^2(\cdot)$ is a stationary, isotropic covariance function with parameters $\boldsymbol\Delta$. The parameters $\boldsymbol\Delta$ as well as B-spline coefficients $\boldsymbol\theta_1$ and $\boldsymbol\theta_2$ are estimated by minimizing \[Q(\boldsymbol\theta_1, \boldsymbol\theta_2, \boldsymbol\Delta) \propto |\boldsymbol\Sigma|^{-n/2} \exp\left\{-\frac{1}{2}\left(\mathbf{Y} - \boldsymbol\mu \right)^t \boldsymbol\Sigma^{-1} \left(\mathbf{Y} - \boldsymbol\mu \right)\right\},\] subject to \[ |\mathbf{J}| = \frac{\partial f_1}{\partial x_1} \times \frac{\partial f_2}{\partial x_2} - \frac{\partial f_1}{\partial x_2} \times \frac{\partial f_2}{\partial x_1} > 0.\] 

## Installation

The most recent version of `bsplinedef` is available at <https://github.com/guiludwig/bsplinedef>. To install the package directly from github, we recommed the `install_github` function available from the `devtools` package. 

```{r eval=FALSE}
library(devtools) # install if missing
install_github("guiludwig/bsplinedef")
```

The `bsplinedef` package depends on `RandomFields`, `fields` and `splines`, among others. It also uses `knitr` and `rmarkdown` to build the documentation.

## Usage: simulated example

First, we will consider a simulated example. 

```{r}
set.seed(1)
n <- 40
x1 <- runif(n)
x2 <- runif(n)
x <- cbind(x1,x2)
F1 <- function(x1, x2) {
  x1 
}
F2 <- function(x1, x2) {
  sqrt(x1*x2) 
}
```

The following figure displays the sampled spatial locations (in black) as well as the deformed locations with coordinates $y_1 = f_1(x_1,x_2)$ and $y_2 = f_2(x_1,x_2)$, in red. The black regular grid is also shown, deformed, in R. The marginal color maps display the functions $f_1$, $f_2.$

```{r echo=FALSE, fig.align="center", message=FALSE, fig.width=6, fig.height=6}
library(bsplinedef)
plotGrid(list(window = list(x = range(x1), y = range(x2)),
              x = x, def.x = cbind(F1(x1,x2), F2(x1,x2))),
         F1 = F1, F2 = F2, margins = TRUE)
```

The `bsplinedef` inherits some syntax from the `RandomFields` package, particularly how to set up a covariance model for the D-distance. First, we will simulate data $Y$ on the D-coordinates $(y_1, y_2)$ using a stationary covariance model, with functions from the `RandomFields` package.

```{r message=FALSE}
# Sets up a Matern covariance model
covModel <- RMexp(var = 1, scale = 1) + RMnugget(var = 1)
data <- RFsimulate(covModel, x = F1(x1,x2), y = F2(x1,x2))
# Extracts values from object RFspatialPointsDataFrame
y  <- as.numeric(unlist(data@data))
```

For this example the spatial locations are simply stored in a $n \times 2$ matrix, while the response values are stored in a $n$-lenght numeric vector.

```{r}
str(x)
str(y)
```

The function `bdef` finds the appropriate deformations $\hat{f}_1,$ $\hat{f}_2,$ as well as estimating the parameters for the spatial covariance function. The default covariance function is a Mat&eacute;rn covariance function plus nugget effect, with differentiability parameter $\nu = 2.5$ (fixed), unknown variance $\sigma_\eta^2$, unknown scale parameter $\gamma$ and unknown nugget effect $\sigma_\varepsilon^2.$ In this case, we need to modify the default action to use an exponential covariance.

For example, fitting the model to the simulated dataset:

```{r}
# Parameters to be estimated set as NA
cov.to.estimate <- RMexp(var = NA, scale = NA) + RMnugget(var = NA)
model <- bdef(x, y, cov.model = cov.to.estimate)
```

Estimated parameters can be recovered with

```{r}
model$model # Covariance model
```

If, for some reason, one needs the spline coefficients, they can also be extracted with

```{r}
model$theta1 # Spline coefficients for f_1
```

The fitting process supports some customization. For example, one may be interested in a different covariance function. We may change it with the following command:

```{r}
# Value set to $nu = 2.5$
G <- RMmatern(nu = 2.5, var = NA, scale = NA) # No nugget
# documentation, values left as NA are to be estimated
model.matern <- bdef(x, y, cov.model = G)
model.matern$model
```

## Plotting tools

We provide two plotting functions along with the `bsplinedef` package. The first one, named `plotGrid`, displays the deformation grid on the data.

```{r fig.align="center", message=FALSE, fig.width=6, fig.height=6}
plotGrid(model, margins = TRUE)
```

The second function, `plotGDdist`, produces a comparison between sample covariances using the G-domain and D-domain distances for reference. It only works if all sampled sites have more than one observation, which might require pruning of the data to use. It still needs to be fully implemented as of `bsplinedef` version `0.1.6` (for example, the `lines` argument, which prints the estimated covariance function decay, is missing).

```{r}
try(plotGDdist(model)) # Fails
```

```{r}
# Simpler data set
RFoptions(Ttriple = FALSE)
n <- 10; t <- 3
x1 <- runif(n)
x2 <- runif(n)
x <- cbind(rep(x1, each = t),
           rep(x2, each = t),
           rep(1:t, n))
covModel <- RMmatern(proj = "space", nu = 2.5, var = 1, scale = 2) +
            RMexp(proj = "time", var = 1, scale = 2)
data <- RFsimulate(covModel, x = F1(x1,x2), 
                             y = F2(x1,x2),
                             T = 1:t)
y  <- as.numeric(unlist(data@data))
```

Fits model

```{r}
G2 <- RMmatern(proj = "space", nu = 2.5, var = NA, scale = NA) + 
      RMexp(proj = "time", var = NA, scale = NA)
# model.temporal <- bdef(x, y, cov.model = G2)
# Not working yet! Need to debug why.
```